---
title: "Portfolio"
author: "Rik de Vries"
date: "2/19/2022"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    self_contained: false
    vertical_layout: fill
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(spotifyr)
library(tidyverse)
library(plotly)
library(flexdashboard)
library(compmus)
all_lp_tracks <- get_playlist_audio_features("", "1y9QYGfYFzSO7EF53wwLiB")
```
### Linkin Park's albums over time as the corpus selection
<center>
<h1>Linkin Park: Old vs New</h1>
<img src="https://i.scdn.co/image/ab67616d0000b273e2f039481babe23658fc719a" alt="Hybrid Theory" width="20%"/>
<img src="https://i.scdn.co/image/ab67616d0000b27391e2fd0e284ca923b8743b6a" alt="Reanimation" width="20%"/>
<img src="https://i.scdn.co/image/ab67616d0000b273ce4b724f4332773136067d8c" alt="Meteora" width="20%"/>

<img src="https://i.scdn.co/image/ab67616d0000b27346e207de66ba06422897f769" alt="Minutes to Midnight" width="20%"/>
<img src="https://i.scdn.co/image/ab67616d0000b273302a793ccd682e6549c91f10" alt="A Thousand Suns" width="20%"/>
<img src="https://i.scdn.co/image/ab67616d0000b273987fb4c5ec8790e9f637a4a4" alt="LIVING THINGS" width="20%"/>

<img src="https://i.scdn.co/image/ab67616d0000b2737cb20e0a29701a667dac3ede" alt="Recharged" width="20%"/>
<img src="https://i.scdn.co/image/ab67616d0000b2735721adac031512b903f10d03" alt="The Hunting Party" width="20%"/>
<img src="https://i.scdn.co/image/ab67616d0000b273145e1cf12538f5666511237d" alt="One More Light" width="20%"/>
</center>

***
<iframe src="https://open.spotify.com/embed/playlist/7MATpbz5p1afYISxOgY7ek?utm_source=generator" width="100%" height="380" frameBorder="0" allowfullscreen="" allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture"></iframe>

For my corpus, I was thinking of selecting different albums from Linkin Park on Spotify. I feel like this would make a very interesting corpus as the band has made a lot of very large changes over the years. Their early albums were very raw, producing a lot of music in the nu metal genre. The type of music they put out later on was much softer, which is something critics and some types of fans did not always like. I would like to identify the differences between these early and late albums.

If possible, I'd also like to compare the band's music to some of the band's side projects as well, like Chester Bennington's Grey Daze and Dead By Sunrise bands and Mike Shinoda's Fort Minor songs along with his new solo career that he started after Chester committed suicide.

A special album that I am very interested in is the Reanimation album. This is one of the top 5 best-selling remix albums, even to this day. Of course, as this only contains remixes, there should be a lot of visible differences to the rest of the band's work.

I feel like comparing all these songs would give a decent picture of how the band and its members changed over time. There are some limitations to this corpus, as the band used to have a different lead singer, Mark Wakefield, but they never released an official album back then. Next to that, the group has released many "Underground" albums, which are not on Spotify. These would have been very interesting to include in the corpus as well, to see if there are visible differences between the officially released albums and the albums that did not receive the same type of polishing.

### Album popularity dips but picks up to original heights
```{r pop_plot2, fig.height = 6, fig.width = 10, fig.align = "center", echo=FALSE}
# Filter out songs that have been duplicated
# (republished under the 20th anniversary album), along with inbetween-tracks
# that are shorter than a minute as these are not songs.
unq_lp_tracks = distinct(all_lp_tracks, track.name, .keep_all = TRUE) %>% 
  filter(track.duration_ms > 60000, track.album.album_type == "album")

unq_lp_tracks$year_album_published = 
  as.Date(format(as.Date(unq_lp_tracks$track.album.release_date, 
                         format="%Y-%m-%d"),"%Y"), "%Y")

main_albums <- c("Hybrid Theory (Bonus Edition)","Meteora (Bonus Edition)", "Minutes to Midnight",  "A Thousand Suns", "LIVING THINGS","The Hunting Party", "One More Light")
main_album_tracks <- unq_lp_tracks %>% filter(track.album.name %in% main_albums)
main_albums_plot <- ggplot(main_album_tracks, aes(x=year_album_published, y=track.popularity)) +
  labs(title="Linkin Park track popularity over time", x="Year", y="Popularity", color = "Album") +
  lims(y=c(0, 100)) +
  geom_point(aes(color=track.album.name, text=sprintf("Track: %s<br>Album: %s<br>Popularity: %s", track.name, track.album.name, track.popularity))) + 
  geom_smooth(aes(text=""))
#"Year: %s<br>Popularity: %s", year_album_published, track.popularity
w <- ggplotly(main_albums_plot, tooltip=c("text"))

# Disable hover for geom_smooth (trace 9) (Lower trace indices are for albums)
w %>%
  style(hoverinfo = "none", traces = 9)
```

***
The following plot shows the popularity of tracks from the main albums of the band Linkin Park. 
Data points on the chart can be hovered to get specific information (track name, the album it belongs to and its popularity).

To give a clear image of the band's popularity, the plot only relies on tracks from main albums. It therefore excludes remix albums and albums made after Chester Bennington's death.
The plot shows a trend that starts off downward sloping, with an all-time low in 2010. After this lowpoint though, it climbs back up to early 2000-levels. 

### Tracks get more well-defined pitch classes over time 
```{r echo=FALSE}
one_more_light <-
  get_tidy_audio_analysis("3xXBsjrbG1xQIm1xv1cKOt") %>%
  select(segments) %>%
  unnest(segments) %>%
  select(start, duration, pitches)
in_the_end <-
  get_tidy_audio_analysis("60a0Rd6pjrkxjPbaKzXjfq") %>%
  select(segments) %>%
  unnest(segments) %>%
  select(start, duration, pitches)
```

```{r fig.height = 6, fig.width = 10, fig.align = "center", echo=FALSE}
plot_oml <- one_more_light %>%
  mutate(pitches = map(pitches, compmus_normalise, "chebyshev")) %>%
  compmus_gather_chroma() %>% 
  ggplot(
    aes(
      x = start + duration / 2,
      width = duration,
      y = pitch_class,
      fill = value,
      text = sprintf("Time: %.0f seconds<br>Pitch class: %s<br>Magnitude: %s", start + duration / 2, pitch_class, value)
    )
  ) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
plot_ite <- in_the_end %>%
  mutate(pitches = map(pitches, compmus_normalise, "chebyshev")) %>%
  compmus_gather_chroma() %>% 
  ggplot(
    aes(
      x = start + duration / 2,
      width = duration,
      y = pitch_class,
      fill = value,
      text = sprintf("Time: %.0f seconds<br>Pitch class: %s<br>Magnitude: %s", start + duration / 2, pitch_class, value)
      )
  ) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()

plt1 <- ggplotly(plot_ite, tooltip="text")
plt2 <- ggplotly(plot_oml, tooltip="text")
w <- subplot(plt1, plt2)
w %>%
  layout(title = 'In the End vs One More Light')

#w %>%
#  
#  #style(hoverinfo = "none") %>%
#  layout(annotations = list(
# list(x = 0.17 , y = 1.065, text = "In the End", showarrow = F, xref='paper', yref='paper'),
#  list(x = 0.87 , y = 1.065, text = "One More Light", showarrow = F, xref='paper', yref='paper'))
#)

```
***
This plot shows the chromagrams from two of Linkin Park's most popular songs, one from their first album and one from their last album. The song on the left is *In the End*, and the song on the right is *One More Light*. 

The first song changes a lot over the course of its duration. It has many instruments: piano, guitar, drums, electric instruments and there seems to be some electric crackling noises added too. On top of all of that, it has both Chester Bennington singing loudly, mixed with Mike Shinoda rapping. This leads to a very mixed chromagram where, even though the main notes are clearly visible, there is a lot of spread to other notes. 

One More Light takes a very different approach. It has very well-defined chroma's with barely any spread to other pitch classes. From the chromagram it can be seen that Chester sings on-key very well. The song is much more emotional, and less rock-like than *In the End*. It does not nearly contain as many instruments, or as much diversity. This displays a large difference between the two songs and their respective albums.

### Older and newer songs have similar structures, but very different execution
```{r echo=FALSE}
# In the end 60a0Rd6pjrkxjPbaKzXjfq
# One more light 3xXBsjrbG1xQIm1xv1cKOt
ite_plot <-
  get_tidy_audio_analysis("60a0Rd6pjrkxjPbaKzXjfq") %>%
  compmus_align(bars, segments) %>%
  select(bars) %>%
  unnest(bars) %>%
  mutate(
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "acentre", norm = "manhattan"
      )
  ) %>%
  mutate(
    timbre =
      map(segments,
        compmus_summarise, timbre,
        method = "mean"
      )
  )
mix_plot_1 <-bind_rows(
  ite_plot %>%
    compmus_self_similarity(pitches, "aitchison") %>%
    mutate(d = d / max(d), type = "Chroma"),
  ite_plot %>%
    compmus_self_similarity(timbre, "euclidean") %>%
    mutate(d = d / max(d), type = "Timbre")
) %>%
  mutate() %>%
  ggplot(
    aes(
      x = xstart + xduration / 2,
      width = xduration,
      y = ystart + yduration / 2,
      height = yduration,
      fill = d
    )
  ) +
  geom_tile() +
  coord_fixed() +
  facet_wrap(~type) +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "")

oml_plot <-
  get_tidy_audio_analysis("3xXBsjrbG1xQIm1xv1cKOt") %>%
  compmus_align(bars, segments) %>%
  select(bars) %>%
  unnest(bars) %>%
  mutate(
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "acentre", norm = "manhattan"
      )
  ) %>%
  mutate(
    timbre =
      map(segments,
        compmus_summarise, timbre,
        method = "mean"
      )
  )
mix_plot_2 <- bind_rows(
  oml_plot %>%
    compmus_self_similarity(pitches, "aitchison") %>%
    mutate(d = d / max(d), type = "Chroma"),
  oml_plot %>%
    compmus_self_similarity(timbre, "euclidean") %>%
    mutate(d = d / max(d), type = "Timbre")
) %>%
  mutate() %>%
  ggplot(
    aes(
      x = xstart + xduration / 2,
      width = xduration,
      y = ystart + yduration / 2,
      height = yduration,
      fill = d
    )
  ) +
  geom_tile() +
  coord_fixed() +
  facet_wrap(~type) +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "")

subplot(mix_plot_1, mix_plot_2, nrows = 2, margin = 0.08) %>%
  layout(annotations = list( 
    list(
      x = 0.1,
      y = 1,
      text = "In the End",  
      xref = "paper",  
      yref = "paper",  
      xanchor = "center",  
      yanchor = "bottom",  
      showarrow = FALSE 
    ),  
    list( 
      x = 0.1,  
      y = 0.42,
      text = "One More Light",  
      xref = "paper",  
      yref = "paper",  
      xanchor = "center",  
      yanchor = "bottom",  
      showarrow = FALSE 
  )))
#ggplotly(mix_plot_1)
#ggplotly(mix_plot_2)

```

***
This plot shows the self-similarity matrices of *In the End* and *One More Light*, and it tells us an important difference between old and new songs. As could be seen in the chromagrams on the previous tab, *In the End* starts off with a D#, F, F# and A#. These are almost nowhere to be seen in the rest of the song, so it is only similar to itself. The song then goes on to play some different notes that persist for almost the whole song, with C# being the most pronounced. 
The timbre self-similarity matrix tells an equally interesting story. The instrumentals in the beginning are definitely unique. That is because they are mixed in with electric/static noises. If they would not have been, they would have been similar to the instrumentals in the outro. The matrix shows the layout of the song very well. At around 20 seconds, Mike Shinoda starts rapping. This is something he also does at around 80 seconds, which is why the self-similarity matrix shows a connection between these two time points. Chester Bennington sings the chrous three times, with the third time being led by a bridge. The three areas are very visible in the matrix. Half of the bridge is sung in a calm way that is unique to any other moment in the song. The other half of the bridge sounds a lot like the chorus, which is why the square that starts at 150 seconds is so much larger than the other two.

*One More Light* is sung in a very different way. A lot of the song is sung in the G# key, as can be seen in the chromagram on the previous tab. This is mixed in with a bit of the F key and some of the C# chroma.
The repetition of these three chromas (mainly the G# key) makes for the pattern we see in the self-similarity matrix. The dark square we see at 70 seconds is the first repetition of the chorus. The chromagram does not clearly show when the chorus repeats though, for this we need to look at the timbre matrix. This clearly shows the chorus is song again at around 150 seconds and then a final time at 210 seconds.
At around 3 minutes in we see a dark blue square that is not similar to anything in the song, except a bit to the intro. This is a relaxed instrumental bridge. It is similar to the intro because the this also only contained instrumental music. The two yellow lines visible at 125 and 135 seconds are long pauses between two sentences, where there is almost no sound at all. These are not similar to any other points in the song.

So while both *In the End* and *One More Light* clearly show verses and choruses being similar, the chromagram shows that *One More Light* alternates a lot more between chromas than *In the End*, which looks like a much more linear song. The timbre matrices show that both songs have unique points next to repetition. In *One More Light* these are slow instrumental pieces, or moments of complete silence. In *In the End* this was a bridge where there was a moment of less intense singing, which then transitioned into the last repetition of the chorus.
<!---
Comparisons with reviewers, 
Style of albums (loudness, dancability, ...)
-->