---
title: "Portfolio"
author: "Rik de Vries"
date: "2/19/2022"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(spotifyr)
library(tidyverse)
library(plotly)
library(flexdashboard)
all_lp_tracks <- get_playlist_audio_features("", "1y9QYGfYFzSO7EF53wwLiB")
```
### Corpus selection
For my corpus, I was thinking of selecting different albums from Linkin Park on Spotify. I feel like this would make a very interesting corpus as the band has made a lot of very large changes over the years. Their early albums were very raw, producing a lot of music in the nu metal genre. The type of music they put out later on was much softer, which is something critics and some types of fans did not always like. I would like to identify the differences between these early and late albums.

If possible, I'd also like to compare the band's music to some of the band's side projects as well, like Chester Bennington's Grey Daze and Dead By Sunrise bands and Mike Shinoda's Fort Minor songs along with his new solo career that he started after Chester committed suicide.

A special album that I am very interested in is the Reanimation album. This is one of the top 5 best-selling remix albums, even to this day. Of course, as this only contains remixes, there should be a lot of visible differences to the rest of the band's work.

I feel like comparing all these songs would give a decent picture of how the band and its members changed over time. There are some limitations to this corpus, as the band used to have a different lead singer, Mark Wakefield, but they never released an official album back then. Next to that, the group has released many "Underground" albums, which are not on Spotify. These would have been very interesting to include in the corpus as well, to see if there are visible differences between the officially released albums and the albums that did not receive the same type of polishing.

### Album popularity
```{r pop_plot2, fig.height = 6, fig.width = 10, fig.align = "center", echo=FALSE}
# Filter out songs that have been duplicated
# (republished under the 20th anniversary album), along with inbetween-tracks
# that are shorter than a minute as these are not songs.
unq_lp_tracks = distinct(all_lp_tracks, track.name, .keep_all = TRUE) %>% 
  filter(track.duration_ms > 60000, track.album.album_type == "album")

unq_lp_tracks$year_album_published = 
  as.Date(format(as.Date(unq_lp_tracks$track.album.release_date, 
                         format="%Y-%m-%d"),"%Y"), "%Y")

main_albums <- c("Hybrid Theory (Bonus Edition)","Meteora (Bonus Edition)", "Minutes to Midnight",  "A Thousand Suns", "LIVING THINGS","The Hunting Party", "One More Light")
main_album_tracks <- unq_lp_tracks %>% filter(track.album.name %in% main_albums)
main_albums_plot <- ggplot(main_album_tracks, aes(x=year_album_published, y=track.popularity)) +
  labs(title="Linkin Park track popularity over time", x="Year", y="Popularity", color = "Album") +
  lims(y=c(0, 100)) +
  geom_point(aes(color=track.album.name, text=sprintf("Track: %s<br>Album: %s<br>Popularity: %s", track.name, track.album.name, track.popularity))) + 
  geom_smooth(aes(text=""))
#"Year: %s<br>Popularity: %s", year_album_published, track.popularity
w <- ggplotly(main_albums_plot, tooltip=c("text"))

# Disable hover for geom_smooth (trace 9) (Lower trace indices are for albums)
w %>%
  style(hoverinfo = "none", traces = 9)
```

***
The following plot shows the popularity of tracks from the main albums of the band Linkin Park. 
Data points on the chart can be hovered to get specific information (track name, the album it belongs to and its popularity).

To give a clear image of the band's popularity, the plot only relies on tracks from main albums. It therefore excludes remix albums and albums made after Chester Bennington's death.
The plot shows a trend that starts off downward sloping, with an all-time low in 2010. After this lowpoint though, it climbs back up to early 2000-levels. 
<!---
Comparisons with reviewers, 
Style of albums (loudness, dancability, ...)
-->